---
title: "HAI code"
author: "Alicia Lowe"
date: "2025-09-29"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Library selection

```{r}
library(readr) # Read in data files
library(dplyr) # Data Cleaning
library(tidyr) # Data Cleaning
library(lubridate) # Change date format for Data cleaning
library(stringr) # Concatenate character groups to allow table collapsing
library(ggformula) # Graphing
library(ggplot2) # Graphing 
library(gridExtra) # Wrap graphs
library(gt) # Create tables
library(gtsummary)  # Create summary tables
library(reshape2) # to create bar graph with proportion of HAI to non-HAI
library(corrplot) # Create Correlation plots
library(FactoMineR) # FAMD analysis
library(factoextra) # Create FAMD variable plot
library(caret) # Cross validation for Model selection
library(themis) # for recipe use in caret
library(pROC) # ROC curve analysis
library(randomForest) # Random Forest
library(xgboost) # XGBoost
library(nnet) #ANN
library(NeuralNetTools) #ANN
```  

Data
```{r}
#Read in dataset Except for labs which is disjointed due to size
patients = read_csv("core_patients.csv")
diagnoses = read_csv("diagnoses.csv.gz")
microbiology = read_csv("microbiology.csv.gz")
procedures = read_csv("procedures.csv.gz")
drugs = read_csv("pharmacy.csv.gz")

```

```{r}
#Look at the classes of the dataset
glimpse(patients)
glimpse(diagnoses)
glimpse(microbiology)
glimpse(procedures)
glimpse(drugs)
```

```{r}
#Read in dataset for labs
labs1 = read_csv("icu_labs_part1.csv.gz")
labs2 = read_csv("icu_labs_part2.csv.gz")
labs3 = read_csv("icu_labs_part3.csv.gz")
labs4 = read_csv("icu_labs_part4.csv.gz")
```

Join all the labs tables together
```{r}
#Merge the 4 labs tables into 1 table
labs <- bind_rows(labs1, labs2, labs3, labs4)
```

###Data Preprocessing and Cleaning
Missing data
```{r, echo=FALSE}
colSums(is.na(patients))
colSums(is.na(diagnoses))
colSums(is.na(microbiology))
colSums(is.na(procedures))
colSums(is.na(drugs))
colSums(is.na(labs))
```

summary of data.
```{r, echo=FALSE}
summary(patients)
summary(diagnoses)
summary(diagnoses)
summary(procedures)
summary(drugs)
summary(labs)
```


Change all characters to lower-case
```{r}
patients <- patients %>%
  mutate(across(c(gender, admission_type, admission_location), ~tolower(.)))

diagnoses <- diagnoses %>%
  mutate(across(c(icd_code, diagnosis_description), ~tolower(.)))

microbiology <- microbiology %>%
  mutate(across(c(spec_type_desc, test_name, org_name, ab_name, interpretation, comments),
                ~ tolower(.)))

procedures <- procedures %>%
  mutate(across(c(icd_code, long_title), ~tolower(.)))

drugs <- drugs %>%
  mutate(across(c(medication, route, frequency), ~tolower(.)))

labs <- labs %>%
  mutate(across(c(value, flag, label, fluid, comments), ~tolower(.)))

```


Append table names for columns for joining with duplicate column names
```{r}
diagnoses <- diagnoses %>%

  rename_with(~ paste0(., "_diag"), 
              .cols = -c(subject_id, hadm_id, stay_id, diagnosis_description))

procedures <- procedures %>%
  rename_with(~ paste0(., "_proc"), 
              .cols = -c(subject_id, hadm_id, stay_id, long_title))

labs <- labs %>%
  rename_with(~ paste0(., "_lab"), 
              .cols = -c(subject_id, hadm_id, stay_id, itemid, value, valuenum, flag,
                         label, fluid))
```

Convert dates to date/time
```{r}
# Convert procedure dates 
procedures <- procedures %>%
  mutate(p_charttime = as.POSIXct(chartdate_proc, format = "%Y-%m-%d", tz = "UTC"))
```

##Address Missing Data
Microbiology Missingness
```{r}
#Change NA in org_id where culture is in the test_name
#AI assisted with test_name containing clostridium difficile
microbiology <- microbiology %>%
  mutate(
    is_culture = grepl("culture", test_name, ignore.case = TRUE),
    is_cdiff = grepl("clostridium difficile|c\\.? *diff", test_name, ignore.case = TRUE),
    is_respiratory = grepl("respiratory|sputum|bronchial", spec_type_desc, ignore.case = TRUE),
    is_catheter = grepl("urine", spec_type_desc, ignore.case = TRUE)
  ) %>%
  filter(is_culture | is_cdiff | is_respiratory | is_catheter) %>%
  mutate(
    org_name = case_when(
      is_culture & is.na(org_name) ~ "no growth",
      !is_culture & is.na(org_name) ~ "not culture",
      TRUE ~ org_name
    )
  )

```

Missing susceptibility results such as ab_itemid, ab_name, interpretation are not necessary for modeling at this time will leave as is. Comments will be left as is.

Pharmacy missingness
```{r}
#Any medication not recorded was changed to none since >5% missing
drugs <- drugs %>%
  mutate(medication = ifelse(is.na(medication), "none", medication))

#change NA values in the route and frequency columns
drugs <- drugs %>%
  mutate(route = ifelse(!is.na(route), "unknown", medication))

drugs <-drugs %>%
  mutate(frequency = ifelse(!is.na(frequency), "unknown", frequency))

```

Labs missingness
```{r}
#test label is really test result field
#Drop all columns where the result field is specimen type
labs <- labs %>%
  filter(!label %in% c("specimen type"))

#remove columns where testing is cancelled
labs <- labs %>%
  filter(
    !(grepl("cancel", value, ignore.case = TRUE) |
      (is.na(value) & grepl("cancel|unable to report|error", comments_lab, ignore.case = TRUE)))
  )

#some values were resulted under comment. Parsed comment to Value
labs <- labs %>%
  mutate(
    # AI assisted
    value = ifelse(
      (is.na(value) | value == "") & !is.na(comments_lab) & comments_lab != "",
      comments_lab,
      value
    )
  )

colSums(is.na(labs))
#This leaves only 856 NA columns, Okay to drop remaining as they are truly NA
labs <- labs %>%
  filter(!is.na(value))

```

Creating the HAI flag
```{r}
#Create the HAI flag
#create a list of procedure ICD codes for urine caths and ventilation
proc_times <- procedures %>%
  mutate(
    is_cath = icd_code_proc %in% c("5794", "5795", "9648"),
    is_vent = icd_code_proc %in% c("9671", "9605", "9672", "9390", "9670",
                              "5a1945z", "5a1955z", "5a1935z", "5a09457",
                              "5a0945a", "5a09357", "5a09557", "5a0955a",
                              "5a0935a", "5a0935z", "5a09358", "5a0945z")
  ) %>%
  select(stay_id, p_charttime, is_cath, is_vent)
#Create the exclusion results
#Organisms related to normal skin contaminates and cancelled tests
exclude_orgs <- c("no growth",
                  "staphylococcus, coagulase negative",
                  "staphylococcus epidermidis", 
                  "streptococcus anginosus",
                  "salmonella dublin",
                  "rothia species",
                  "rothia mucilaginosa",
                  "additional information to follow",
                  "negative",
                  "not culture")
#Join patient table with micro table to calculate time from admission
#Join procedure times to micro table to calculate time to test from procedure
micro_relevant <- microbiology %>%
  left_join(patients %>% select(stay_id, intime), by = "stay_id") %>%
  left_join(proc_times, by = "stay_id") %>%
  mutate(
    time_from_icu_admit = as.numeric(difftime(charttime, intime, units = "hours")),
    time_from_proc = as.numeric(difftime(charttime, p_charttime, units = "hours"))
  )

#Initialize the flag as "not HAI"
micro_relevant <- micro_relevant %>%
  mutate(HAI_flag = "not_hai")

#Flag positive blood cultures
blood_idx <- micro_relevant$spec_type_desc %in% 
               c("blood culture", "blood culture ( myco/f lytic bottle)") &
             micro_relevant$time_from_icu_admit > 48 &
             !is.na(micro_relevant$org_name) &
             !grepl("cancelled", micro_relevant$org_name) &
             !(micro_relevant$org_name %in% exclude_orgs)
micro_relevant$HAI_flag[blood_idx] <- "hai"

#Flag C. Diff tests
cdiff_idx <- micro_relevant$is_cdiff &
             micro_relevant$time_from_icu_admit > 48 &
             !is.na(micro_relevant$org_name) &
             micro_relevant$org_name %in% c("clostridium difficile", "positive")
micro_relevant$HAI_flag[cdiff_idx] <- "hai"

#Flag urine catheter related tests
cath_idx <- micro_relevant$is_cath &
            grepl("urine", micro_relevant$spec_type_desc, ignore.case = TRUE) &
            micro_relevant$time_from_proc > 72 &
            !is.na(micro_relevant$org_name) &
            !(micro_relevant$org_name %in% exclude_orgs)
micro_relevant$HAI_flag[cath_idx] <- "hai"

#Flag HAI ventilator related tests
resp_idx <- micro_relevant$is_vent &
            grepl("respiratory|sputum|bronchial", micro_relevant$spec_type_desc, ignore.case = TRUE) &
            micro_relevant$time_from_proc > 72 &
            !is.na(micro_relevant$org_name) &
            !(micro_relevant$org_name %in% exclude_orgs)
micro_relevant$HAI_flag[resp_idx] <- "hai"

#change HAI_flag to factor
micro_relevant <- micro_relevant %>%
  mutate(
    HAI_flag = ifelse(HAI_flag == "hai" & grepl("cancelled", org_name), "not_hai", HAI_flag),
    HAI_flag = factor(HAI_flag, levels = c("not_hai", "hai"))
  )

```

Create adjusted length of stay for patients with HAIs. HAIs would most likely prolong a patient's stay in the ICU therefore adding additional weight in modeling for icu_los falsely elevating the importance of the predictors icu_los
```{r}
hai_times <- micro_relevant %>%
  filter(HAI_flag == "hai") %>%
  group_by(stay_id) %>%
  summarise(first_hai_time = min(charttime, na.rm = TRUE), .groups = "drop")

# Join with patient data
patients <- patients %>%
  left_join(hai_times, by = "stay_id") %>%
  mutate(
    icu_los = as.numeric(difftime(outtime, intime, units = "hours")),
    adjusted_icu_los = case_when(
      !is.na(first_hai_time) ~ as.numeric(difftime(first_hai_time, intime, 
                                                   units = "hours")),
      TRUE ~ icu_los
    )
  )
```

Remove additional rows into 1 row per patient stay Results
```{r}
# Collapse microbiology table to patient-level HAI summary
micro_patient <- micro_relevant %>%
  group_by(stay_id, subject_id, hadm_id) %>%
  summarise(
    # AI assisted
    chartdate = paste(unique(chartdate), collapse = "; "),
    charttime = paste(unique(format(charttime, "%Y-%m-%d %H:%M:%S")), collapse = "; "),
    spec_type_desc = paste(unique(spec_type_desc), collapse = "; "),
    test_name = paste(unique(test_name), collapse = "; "),
    org_name = paste(unique(org_name), collapse = "; "),
    HAI_flag = ifelse(any(HAI_flag == "hai"), "hai", "not_hai"),
    cath_HAI = any(is_cath, na.rm = TRUE),
    respiratory_HAI = any(is_vent, na.rm = TRUE),
    
    .groups = "drop"
  )
```

Check positivity of HAI flag
```{r}
#Calculating the HAI positivity rate from collapsed table.
micro_patient %>%
  count(HAI_flag) %>%
  mutate(percent = n / sum(n) * 100)

```

Join Patient table to micro_patient table
```{r}
# Join patient-level features to your stay-level microbiology summary
micro_patient <- micro_patient %>%
  left_join(
    patients %>% 
      select(subject_id, gender, anchor_age, admission_type, admission_location, stay_id,
             intime, outtime, icu_los, adjusted_icu_los),
    by = c("stay_id", "subject_id")
  )
```

Collapse the Diagnoses table per stay_id
```{r}
#Collapse diagnosis by stay
#Create flags based on common illness associated with increase infection risks 
diagnoses_stay <- diagnoses %>%
  group_by(stay_id, subject_id, hadm_id) %>%
  summarise(
    # AI assisted
    icd_codes = paste(unique(icd_code_diag), collapse = "; "),
    diagnoses_text = paste(unique(diagnosis_description), collapse = "; "),
    num_diagnoses = n(),
    has_diabetes = any(str_detect(diagnosis_description, 
                                  "diabetes|hyperglycemia|hypoglycemia")),
    has_ckd = any(str_detect(diagnosis_description, 
                             "chronic kidney|renal failure|
                             acute kidney failure|nephropathy")),
    has_resp = any(str_detect(diagnosis_description, 
                              "chronic respiratory|copd|asthma|emphysema")),
    has_ld = any(str_detect(diagnosis_description, 
                            "liver|hepatitis|cirrhosis|hepatic|biliary")),
    has_cancer = any(str_detect(diagnosis_description, 
                                "cancer|carcinoma|malignant|neoplasm|lymphoma|
                                leukemia|metastatic|tumor")),
    has_hd = any(str_detect(diagnosis_description, 
                            "heart|cardiac|myocardial|coronary|
                            atrial fibrillation|hypertension|ischemic|
                            angina|heart failure")),
    has_hyperlipidemia = any(str_detect(diagnosis_description,
                                        "hyperlipidemia|hypercholesterolemia|
                                        hypertriglyceridemia")),
    has_acute_anemia = any(str_detect(diagnosis_description, 
                                  "acute posthemorrhagic anemia|
                                  acute blood loss anemia")),
    has_acidosis = any(str_detect(diagnosis_description, 
                              "acidosis|lactic acidosis|metabolic acidosis|
                              respiratory acidosis")),
    has_nicotine_depend = any(str_detect(diagnosis_description, "nicotine")),
    has_thrombocytopenia = any(str_detect(diagnosis_description, 
                                      "thrombocytopenia|low platelet",))
    
                                    
  ) %>%
  ungroup()

```

Join diagnoses to micro_patient
```{r}
#Join to master table micro_patient
micro_patient <- micro_patient %>%
  left_join(diagnoses_stay, by = c("stay_id", "subject_id", "hadm_id"))
```

Add flag for missing diagnosis information
```{r}
colSums(is.na(micro_patient))
micro_patient <- micro_patient %>%
  mutate(
    has_diabetes = ifelse(is.na(has_diabetes), 0, has_diabetes),
    has_ckd = ifelse(is.na(has_ckd), 0, has_ckd),
    has_resp = ifelse(is.na(has_resp), 0, has_resp),
    has_ld = ifelse(is.na(has_ld), 0, has_ld),
    has_cancer = ifelse(is.na(has_cancer), 0, has_cancer),
    has_hd = ifelse(is.na(has_hd), 0, has_hd),
    has_hyperlipidemia = ifelse(is.na(has_hyperlipidemia), 0, has_hyperlipidemia),
    has_acute_anemia = ifelse(is.na(has_acute_anemia), 0, has_acute_anemia),
    has_acidosis = ifelse(is.na(has_acidosis), 0, has_acidosis),
    has_nicotine_depend = ifelse(is.na(has_nicotine_depend), 0, has_nicotine_depend),
    has_thrombocytopenia = ifelse(is.na(has_thrombocytopenia), 0, has_thrombocytopenia),
    num_diagnoses = ifelse(is.na(num_diagnoses), 0, num_diagnoses)
  )
```

Only get procedures which occur prior to the infection
```{r}
hai_times <- micro_relevant %>%
  filter(HAI_flag == "hai") %>%
  group_by(stay_id) %>%
  summarise(first_hai = min(charttime, na.rm = TRUE), .groups = "drop")

procedures_hai <- procedures %>%
  left_join(hai_times, by = "stay_id") %>%
  mutate(before_hai = is.na(first_hai) | p_charttime < first_hai) %>%
  filter(before_hai)
```

collapsing Procedures table
```{r}
#Collapse procedures table by stay and add appropriate flags of most common procedures
procedures_stay <- procedures_hai %>%
  mutate(
    has_cath_urn = grepl("urinary catheter|foley|catheter", long_title, ignore.case = TRUE) | 
                   icd_code_proc %in% c("5794", "5795", "9648"),
    
    has_cath_bld = grepl("venous catheter|central venous|superior vena cava", long_title, ignore.case = TRUE) |
                   icd_code_proc %in% c("3893", "3897", "02hv33z", "9672", "3895"),
    
    has_vent = grepl("mechanical ventilation|intubation|ventilator
                     |endotracheal|tracheostomy", long_title, ignore.case = TRUE) |
               icd_code_proc %in% c("9671", "9605", "9672", "9390", "9670",
                               "5a1945z", "5a1955z", "5a1935z", "5a09457",
                               "5a0945a", "5a09357", "5a09557", "5a0955a",
                               "5a0935a", "5a0935z", "5a09358", "5a0945z",
                               "0bh17ez", "311", "0dh63uz"),
    has_feedingtube = grepl("nutritional substance | enteral infusion", long_title,
                            ignore.case = TRUE) |
      icd_code_proc %in% c("966", "9915", "3e0g76z", "3e0g86z"),
    has_bronch_biopsy = grepl("bronchus biopsy", long_title, ignore.case = TRUE) |
      icd_code_proc %in% c("3324"),
    
    has_surg = grepl("surgery|operative|incision|excision|repair|transplant|biopsy|feeding|nutritional|infusion", long_title, ignore.case = TRUE) |
               icd_code_proc %in% c("3e0g76z", "966", "3324", "0bj08zz", "0dh63uz",
                                    "3995")
  ) %>%
  group_by(stay_id, subject_id, hadm_id) %>%
  summarise(
    has_cath_urn = as.integer(any(has_cath_urn)),
    has_cath_bld = as.integer(any(has_cath_bld)),
    has_vent = as.integer(any(has_vent)),
    has_surg = as.integer(any(has_surg)),
    has_feedingtube = as.integer(any(has_feedingtube)),
    has_bronch_biopsy = as.integer(any(has_bronch_biopsy)),
    num_procedures = n(),
    first_procedure = min(p_charttime, na.rm = TRUE),
    .groups = "drop"
  )
```

Join procedures stay level to master micro_patient table
```{r}
micro_patient <- micro_patient %>%
  left_join(procedures_stay, by = c("stay_id", "subject_id", "hadm_id"))
```

Handle NAs not included in the procedures flags
```{r}
micro_patient <- micro_patient %>%
  mutate(
    has_cath_urn       = ifelse(is.na(has_cath_urn), 0, has_cath_urn),
    has_cath_bld       = ifelse(is.na(has_cath_bld), 0, has_cath_bld),
    has_vent           = ifelse(is.na(has_vent), 0, has_vent),
    has_surg           = ifelse(is.na(has_surg), 0, has_surg),
    has_feedingtube    = ifelse(is.na(has_feedingtube), 0, has_feedingtube),
    has_bronch_biopsy  = ifelse(is.na(has_bronch_biopsy), 0, has_bronch_biopsy),
    num_procedures     = ifelse(is.na(num_procedures), 0, num_procedures),
    first_procedure = ifelse(is.na(first_procedure), NA, first_procedure)
  )
colSums(is.na(micro_patient))
```

Create Drug classes
```{r}
#Creating drug classes
drugs <- drugs %>%
  mutate(drug_group = case_when(
    medication %in% c("insulin") ~ "insulin",
    medication %in% c("glucagon") ~ "glucagon",
    grepl("furosemide|chlorothiazide", medication, ignore.case = TRUE) ~ "diuretic",
    grepl("chlorhexidine", medication, ignore.case = TRUE) ~ "antiseptic",
    grepl("vaccine|vac", medication, ignore.case = TRUE) ~ "vaccine",
    grepl("albumin", medication, ignore.case = TRUE) ~ "volume_expander",
    grepl("potassium|magnesium|calcium|sodium|dextrose|glucose|neutra-phos|phosphate|
          phosphorus", medication, ignore.case = TRUE) ~ "electrolytes_fluid",
    grepl("metoprolol|amiodarone|aspirin|heparin(?!flush)|
          warfarin|labetalol|hydralazine|atorvastatin|nitroglycerin|diltiazem|
          vasopressin|amlodipine|nicardipine|clopidogrel|simvastatin|captopril|digoxin|
          carvedilol", medication, ignore.case = TRUE, perl = TRUE) ~ "cardiovascular",
    grepl("vancomycin|cefepime|piperacillin-tazobactam|ceftriaxone|cefazolin|meropenem|
          azithromycin|metronidazole|ciprofloxacin|neomycin|polymyxin|bacitracin|
          fluconazole|levofloxacin|ceftazidime|acyclovir", 
          medication, ignore.case = TRUE) ~ "antibiotic",
    grepl("acetaminophen|fentanyl|hydromorphone|morphine|oxycodone|ketorolac", medication,
          ignore.case = TRUE) ~ "analgesic",
    grepl("propofol|midazolam|lorazepam|dexmedetomidine|diazepam|ramelteon",
          medication, ignore.case = TRUE) ~ "sedative",
    grepl("quetiapine fumarate|haloperidol|trazodone|olanzapine", medication, 
          ignore.case = TRUE) ~ "neuroleptics",
    grepl("senna|docusate|bisacodyl|pantoprazole|famotidine|ondansetron|
          polyethylene glycol|ranitidine|omeprazole|metoclopramide|bicarbonate|
          milk of magnesia|lansoprazole", medication, ignore.case = TRUE) ~ "gi",
    grepl("albuterol|ipratropium|phenylephrine|glycopyrrolate|acetylcysteine", 
          medication, ignore.case = TRUE) ~ "respiratory",
    grepl("levothyroxine", medication, ignore.case = TRUE) ~ "endocrine",
    grepl("thiamine|folic acid|phytonadione|multivitamins|vitamin", medication,
          ignore.case = TRUE) ~ "vitamin",
    grepl("prednisone|methylprednisolone|hydrocortisone", medication, ignore.case = TRUE) 
    ~ "steroid",
    grepl("midodrine|epinephrine", medication, ignore.case = TRUE) ~ "vasopressors",
    grepl("tacrolimus", medication, ignore.case = TRUE) ~ "immunosuppressants",
    grepl("cisatracurium", medication, ignore.case = TRUE) ~ "neuromuscular",
    grepl("diphenhydramine", medication, ignore.case = TRUE) ~ "antihistamine",
    grepl("Heparin Flush|heparin dwell|lidocaine", medication, 
          ignore.case = TRUE) ~ "non_therapeutic",
    grepl("gabapentin", medication, ignore.case = TRUE) ~ "neurologic",
    grepl("sevelamer|calcium carbonate", medication, ignore.case = 
            TRUE) ~ "phosphate_binder",
    grepl("none", medication, ignore.case = TRUE) ~ "none",
    TRUE ~ "other"
  ))
```

Find all prescriptions administered prior to HAI flag
```{r}
drugs_hai <- drugs %>%
  left_join(hai_times, by = "stay_id") %>%
  filter(is.na(first_hai) | starttime < first_hai)
```

Collapsing the Drug table
```{r}
#Group by stay and include drug groups
drugs_stay <- drugs_hai %>% 
  group_by(subject_id, hadm_id, stay_id) %>% 
  summarise( medications = paste(unique(medication), collapse = ", "), 
             drug_groups = paste(unique(drug_group), collapse = ", "),
             .groups = "drop" ) %>% 
  left_join( drugs_hai %>% 
               distinct(subject_id, hadm_id, stay_id, drug_group) %>% 
               mutate(flag = 1) %>% 
               pivot_wider( names_from = drug_group, values_from = flag, 
                            values_fill = 0 ), by = c("subject_id", "hadm_id", "stay_id") )
```

Join drug to micro_patient table
```{r}
#Take care of binary flag NAs by creating a list of all the flags
drug_flags <- c("insulin", "glucagon", "diuretic", "antiseptic","vaccine", "volume_expander","electrolytes_fluid","cardiovascular", "antibiotic","analgesic", "sedative", "neuroleptics", "gi", "respiratory", "endocrine", "vitamin", "steroid", "vasopressors", "immunosuppressants", "neuromuscular", "antihistamine", "non_therapeutic", "neurologic", "phosphate_binder", "none", "other")

micro_patient <- micro_patient %>%
  left_join(drugs_stay, by = c("stay_id", "subject_id", "hadm_id"))

micro_patient <- micro_patient %>%
  mutate(across(any_of(drug_flags), ~replace_na(.x, 0)))

micro_patient <- micro_patient %>%
  mutate(drug_groups = ifelse(is.na(drug_groups), "none", drug_groups))
```

Handling NAs introduced after the join
```{r}
micro_patient <- micro_patient %>%
  mutate(
    medications = ifelse(is.na(medications), "none", medications)
  )
```

Memory issues with labs being so large, remove unnecessary test to only the subset
```{r}
labs <- labs %>%
  filter(label %in% c(
    "glucose", "sodium", "potassium", "ph", "chloride",
    "hemoglobin", "bicarbonate", "creatinine"
  ))
```

##Creating flags for labs
Clean up Glucose tests
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(value = if_else(label == "glucose",
                    str_replace_all(value, "verified|qns to verify|greater than|
                                    less than", ""),value),
    value = if_else(label == "glucose", str_replace_all(value, "[><*]", ""), value),
    value = if_else(label == "glucose", str_replace_all(value, "\\.{3,}", " "), value),
    value = if_else(label == "glucose", str_replace_all(value, "\\s*\\.$", ""), value),
    value = if_else(label == "glucose", str_replace_all(value, "^\\.(?=\\d)", ""), value),
    value = if_else(label == "glucose", str_squish(value), value),
    value = if_else(label == "glucose" & fluid == "urine" & value %in% 
                      c("neg", "neg.", "negative", "n"), "0", value),
    value_num = if_else(label == "glucose", suppressWarnings(as.numeric(value)),NA_real_),
    glucose_ordered = if_else(label == "glucose", 1, 0),
    low_glucose = case_when(
      label == "glucose" & fluid == "blood" & !is.na(value_num) & value_num < 70 ~ 1,
      TRUE ~ 0
    ),
    high_glucose = case_when(
      label == "glucose" & fluid == "blood" & !is.na(value_num) & value_num > 100 ~ 1,
      label == "glucose" & fluid == "urine" & (!is.na(value_num) & value_num > 100) ~ 1,
      label == "glucose" & fluid == "urine" &
        str_detect(value, regex("pos", ignore_case = TRUE)) ~ 1,
      TRUE ~ 0
    )
  )

```

Clean up Sodium Test
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(
    value = case_when(
      label == "sodium" & value %in% c("___", "anion gap verified.") ~ NA_character_,
      TRUE ~ value
    ),
    value = if_else(label == "sodium", str_replace_all(value, "[><*]", ""), value),
    value = if_else(label == "sodium", str_trim(value), value),
    value_num = if_else(label == "sodium",
                        suppressWarnings(as.numeric(value)),
                        NA_real_),
    low_sodium  = if_else(label == "sodium" & !is.na(value_num) & value_num < 135, 1, 0),
    high_sodium = if_else(label == "sodium" & !is.na(value_num) & value_num > 145, 1, 0),
    sodium_ordered   = if_else(label == "sodium", 1, 0)
  )
```

Clean up potassium test results
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(
    value = case_when(
      label == "potassium" & value %in% c("___", "verified by alternate methodology.") 
      ~ NA_character_, TRUE ~ value),
    value = if_else(label == "potassium", str_extract(value, "[><]?\\s*\\d+(\\.\\d+)?"),
                    value),
    value = if_else(label == "potassium", str_replace_all(value, "[><*]", ""), value),
    value = if_else(label == "potassium", str_squish(value), value),
    value_num = if_else(label == "potassium", suppressWarnings(as.numeric(value)),
                        NA_real_),
    low_potassium  = if_else(label == "potassium" & !is.na(value_num) & value_num < 3.6,
                             1, 0),
    high_potassium = if_else(label == "potassium" & !is.na(value_num) & value_num > 5.2,
                             1, 0),
    potassium_ordered = if_else(label == "potassium", 1, 0)
  )

```

Clean up pH test results
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(
    value = case_when(
      label == "ph" & value %in% c("verified.", "notdone", "___") ~ NA_character_,
      TRUE ~ value
    ),
    value = if_else(label == "ph",
                    str_replace_all(value, "verified|greater than|less than", ""),
                    value),
    value = if_else(label == "ph", str_replace_all(value, "[><*]", ""), value),
    value = if_else(label == "ph", str_replace_all(value, "\\.{3,}", " "), value),
    value = if_else(label == "ph", str_squish(value), value),
    value_num = if_else(label == "ph",
                        suppressWarnings(as.numeric(value)),
                        NA_real_),
    ph_ordered = if_else(label == "ph", 1, 0),
    low_ph_bld  = if_else(label == "ph" & fluid == "blood" & !is.na(value_num) & 
                            value_num < 7.3, 1, 0),
    high_ph_bld = if_else(label == "ph" & fluid == "blood" & !is.na(value_num) & 
                            value_num > 7.5, 1, 0),
    low_ph_urn  = if_else(label == "ph" & fluid == "urine"  & !is.na(value_num) &
                            value_num < 4.5, 1, 0),
    high_ph_urn = if_else(label == "ph" & fluid == "urine"  & !is.na(value_num) &
                            value_num > 8.0, 1, 0)
  )

```

Clean up chloride Test
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(
    value = case_when(
      label == "chloride" & value %in% c("___", "anion gap verified.") ~ NA_character_,
      TRUE ~ value
    ),
    value = if_else(label == "chloride", str_replace_all(value, "[><*]", ""), value),
    value = if_else(label == "chloride", str_trim(value), value),
    value_num = if_else(label == "chloride",
                        suppressWarnings(as.numeric(value)),
                        NA_real_),
    low_chloride  = if_else(label == "chloride" & !is.na(value_num) & value_num < 98, 1, 0),
    high_chloride = if_else(label == "chloride" & !is.na(value_num) & value_num > 107, 1, 0),
    chloride_ordered   = if_else(label == "chloride", 1, 0)
  )
```

Clean up hemoglobin results
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(value = case_when(label == "hemoglobin" & value %in% 
                             c("___", "unable to reprot.", 
                               "unable. unable to perform test.", "137.0", "96.0", "98.0")
                           ~ NA_character_, TRUE ~ value),
    value = if_else(label == "hemoglobin",str_extract(value, "\\d+(\\.\\d+)?"), value),
    value = if_else(label == "hemoglobin", str_squish(value), value),
    value_num = if_else(label == "hemoglobin", suppressWarnings(as.numeric(value)),
                        NA_real_),
    hemoglobin_ordered = if_else(label == "hemoglobin", 1, 0)) %>%
  left_join(micro_patient %>% select(stay_id, gender), by = "stay_id") %>%
  mutate(
    low_hemoglobin = case_when(
      label == "hemoglobin" & gender == "m" & !is.na(value_num) & value_num < 13.2 ~ 1,
      label == "hemoglobin" & gender == "f" & !is.na(value_num) & value_num < 11.6 ~ 1,
      label == "hemoglobin" & !is.na(value_num) ~ 0,
      TRUE ~ NA_real_
    ),
    high_hemoglobin = case_when(
      label == "hemoglobin" & gender == "m" & !is.na(value_num) & value_num > 16.6 ~ 1,
      label == "hemoglobin" & gender == "f" & !is.na(value_num) & value_num > 15.0 ~ 1,
      label == "hemoglobin" & !is.na(value_num) ~ 0,
      TRUE ~ NA_real_
    )
  )

```

Clean up bicarbonate Test
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(
    value = case_when(
      label == "bicarbonate" & value %in% c("___") ~ NA_character_,
      label == "bicarbonate" ~ str_replace_all(value, "[><*]", ""), 
      TRUE ~ value  
    ),
    value = if_else(label == "bicarbonate", str_trim(value), value),
    value_num = if_else(label == "bicarbonate",
                        suppressWarnings(as.numeric(value)),
                        NA_real_),
    low_bicarbonate  = if_else(label == "bicarbonate" & !is.na(value_num) & value_num 
                               < 22, 1, 0),
    high_bicarbonate = if_else(label == "bicarbonate" & !is.na(value_num) & value_num 
                               > 29, 1, 0),
    bicarbonate_ordered = if_else(label == "bicarbonate", 1, 0)
  )

```

clean up creatinine test results
```{r}
#Remove meaning less characters, discard illogical results, add flags for abnormal results.
labs <- labs %>%
  mutate(value = case_when(label == "creatinine" & value %in% c("___") ~ NA_character_,
                           TRUE ~ value),
    value = if_else(label == "creatinine",str_replace_all(value, "[><*#]", ""), value),
    value = if_else(label == "creatinine",str_replace(value, "\\.$", ""), value),
    value = if_else(label == "creatinine", str_squish(value), value),
    value_num = if_else(label == "creatinine", suppressWarnings(as.numeric(value)),
                        NA_real_),
    creatinine_ordered = if_else(label == "creatinine", 1, 0)) %>%
  mutate(
    low_creatinine = case_when(
      label == "creatinine" & gender == "m" & !is.na(value_num) & value_num < 0.74 ~ 1,
      label == "creatinine" & gender == "f" & !is.na(value_num) & value_num < 0.59 ~ 1,
      label == "creatinine" & !is.na(value_num) ~ 0,
      TRUE ~ NA_real_
    ),
    high_creatinine = case_when(
      label == "creatinine" & gender == "m" & !is.na(value_num) & value_num > 1.35 ~ 1,
      label == "creatinine" & gender == "f" & !is.na(value_num) & value_num > 1.04 ~ 1,
      label == "creatinine" & !is.na(value_num) ~ 0,
      TRUE ~ NA_real_
    )
  )
```

Find only the test ordered prior to HAI 
```{r}
#Only keeps the tests of interest
select_tests <- c("glucose", "sodium", "potassium", "ph", "chloride", "hemoglobin",
                  "bicarbonate", "creatinine")

labs_col <- labs %>%
  filter(label %in% select_tests)

#Find timing for test before HAI
labs_hai <- labs_col %>%
  left_join(hai_times, by = "stay_id") %>%
  filter(is.na(first_hai) | charttime_lab < first_hai)
```

Collapse labs table
```{r}
#Collapse table with only the tests prior to hai
labs_stay <- labs_hai %>%
  group_by(stay_id, subject_id, hadm_id) %>%
  summarise(
    # Glucose
    low_glucose  = as.integer(any(low_glucose == 1, na.rm = TRUE)),
    high_glucose = as.integer(any(high_glucose == 1, na.rm = TRUE)),
    glucose_ordered   = as.integer(any(glucose_ordered == 1, na.rm = TRUE)),

    # Sodium
    low_sodium  = as.integer(any(low_sodium == 1, na.rm = TRUE)),
    high_sodium = as.integer(any(high_sodium == 1, na.rm = TRUE)),
    sodium_ordered   = as.integer(any(sodium_ordered == 1, na.rm = TRUE)),
    
    # Potassium
    low_potassium  = as.integer(any(low_potassium == 1, na.rm = TRUE)),
    high_potassium = as.integer(any(high_potassium == 1, na.rm = TRUE)),
    potassium_ordered   = as.integer(any(potassium_ordered == 1, na.rm = TRUE)),
    
    # pH
    low_ph_bld      = as.integer(any(low_ph_bld == 1, na.rm = TRUE)),
    high_ph_bld     = as.integer(any(high_ph_bld == 1, na.rm = TRUE)),
    low_ph_urn      = as.integer(any(low_ph_urn == 1, na.rm = TRUE)),
    high_ph_urn     = as.integer(any(high_ph_urn == 1, na.rm = TRUE)),
    ph_ordered      = as.integer(any(ph_ordered == 1, na.rm = TRUE)),
    
    # chloride
    low_chloride  = as.integer(any(low_chloride == 1, na.rm = TRUE)),
    high_chloride = as.integer(any(high_chloride == 1, na.rm = TRUE)),
    chloride_ordered   = as.integer(any(chloride_ordered == 1, na.rm = TRUE)),
    
    # hemoglobin
    low_hemoglobin = as.integer(any(low_hemoglobin == 1, na.rm = TRUE)),
    high_hemoglobin = as.integer(any(high_hemoglobin == 1, na.rm = TRUE)),
    hemoglobin_ordered = as.integer(any(hemoglobin_ordered == 1, na.rm = TRUE)),
    
    # bicarbonate
    low_bicarbonate = as.integer(any(low_bicarbonate == 1, na.rm = TRUE)),
    high_bicarbonate = as.integer(any(high_bicarbonate == 1, na.rm = TRUE)),
    bicarbonate_ordered = as.integer(any(bicarbonate_ordered == 1, na.rm = TRUE)),
   
     # creatinine
    low_creatinine = as.integer(any(low_creatinine == 1, na.rm = TRUE)),
    high_creatinine = as.integer(any(high_creatinine == 1, na.rm = TRUE)),
    creatinine_ordered = as.integer(any(creatinine_ordered == 1, na.rm = TRUE)),
    
    .groups = "drop"
  )
```

Join labs to micro_patient table
```{r}
micro_patient <- micro_patient %>%
  left_join(labs_stay, by = c("stay_id", "subject_id", "hadm_id"))
```

Handle NAs introduced from the join
```{r}
lab_flags <- c(
  "low_glucose", "high_glucose", "glucose_ordered",
  "low_sodium", "high_sodium", "sodium_ordered",
  "low_potassium", "high_potassium", "potassium_ordered",
  "low_ph_bld", "high_ph_bld", "low_ph_urn", "high_ph_urn", "ph_ordered", 
  "low_chloride", "high_chloride", "chloride_ordered",
  "low_hemoglobin", "high_hemoglobin", "hemoglobin_ordered",
  "low_bicarbonate", "high_bicarbonate", "bicarbonate_ordered",
  "low_creatinine", "high_creatinine", "creatinine_ordered"
)

micro_patient <- micro_patient %>%
  mutate(across(all_of(lab_flags), ~ ifelse(is.na(.x), 0, .x)))

```

Drop unnecessary variables
```{r}
cols_to_drop <- c(
  "subject_id", "hadm_id", "chartdate", "charttime", "spec_type_desc", "test_name", 
  "org_name","intime", "outtime", "icd_codes", "diagnoses_text", "first_procedure", 
  "medications", "has_vent", "has_cath_urn",  "drug_groups", "respiratory_HAI", "stay_id",
  "cath_HAI", "icu_los"
  )

clean_patient <- micro_patient %>%
  select(-all_of(cols_to_drop))

```

Convert categorical variables to factors
```{r}
#Change binary HAI flag
clean_patient$HAI_flag <- factor(clean_patient$HAI_flag,
                                 levels = c("hai", "not_hai"),
                                 labels = c("hai", "not_hai"))
clean_patient$HAI_flag <- relevel(clean_patient$HAI_flag, ref = "hai")

clean_patient <- clean_patient %>%
  mutate(
      gender = as.factor(gender), 
      admission_type = as.factor(admission_type), 
      admission_location = as.factor(admission_location),
      has_diabetes = as.factor(has_diabetes), 
      has_ckd = as.factor(has_ckd), 
      has_resp = as.factor(has_resp), 
      has_ld = as.factor(has_ld), 
      has_cancer = as.factor(has_cancer), 
      has_hd = as.factor(has_hd),
      has_cath_bld = as.factor(has_cath_bld), 
      has_surg = as.factor(has_surg), 
      has_feedingtube = as.factor(has_feedingtube), 
      has_thrombocytopenia = as.factor(has_thrombocytopenia),
      has_bronch_biopsy = as.factor(has_bronch_biopsy), 
      has_acidosis = as.factor(has_acidosis), 
      has_hyperlipidemia = as.factor(has_hyperlipidemia), 
      has_acute_anemia = as.factor(has_acute_anemia),
      has_nicotine_depend = as.factor(has_nicotine_depend), 
      cardiovascular = as.factor(cardiovascular),
      other = as.factor(other), gi = as.factor(gi), 
      respiratory = as.factor(respiratory), 
      diuretic = as.factor(diuretic), 
      electrolytes_fluid = as.factor(electrolytes_fluid),
      steroid = as.factor(steroid), 
      sedative = as.factor(sedative), 
      non_therapeutic = as.factor(non_therapeutic), 
      antibiotic = as.factor(antibiotic),
      none = as.factor(none), 
      vitamin = as.factor(vitamin), 
      insulin = as.factor(insulin), 
      analgesic = as.factor(analgesic), 
      vaccine = as.factor(vaccine), 
      neuroleptics = as.factor(neuroleptics),
      antiseptic = as.factor(antiseptic), 
      glucagon = as.factor(glucagon), 
      neurologic = as.factor(neurologic), 
      volume_expander = as.factor(volume_expander),
      antihistamine = as.factor(antihistamine), 
      neuromuscular = as.factor(neuromuscular), 
      vasopressors = as.factor(vasopressors), 
      immunosuppressants = as.factor(immunosuppressants),
      endocrine = as.factor(endocrine), 
      phosphate_binder = as.factor(phosphate_binder), 
      low_glucose = as.factor(low_glucose), 
      high_glucose = as.factor(high_glucose), 
      glucose_ordered = as.factor(glucose_ordered),
      low_sodium = as.factor(low_sodium), 
      high_sodium = as.factor(high_sodium), 
      sodium_ordered = as.factor(sodium_ordered),
      low_potassium = as.factor(low_potassium), 
      high_potassium = as.factor(high_potassium), 
      potassium_ordered = as.factor(potassium_ordered),
      low_ph_bld = as.factor(low_ph_bld), 
      high_ph_bld = as.factor(high_ph_bld), 
      low_ph_urn = as.factor(low_ph_urn), 
      high_ph_urn = as.factor(high_ph_urn), 
      ph_ordered = as.factor(ph_ordered),
      low_chloride = as.factor(low_chloride), 
      high_chloride = as.factor(high_chloride), 
      chloride_ordered = as.factor(chloride_ordered),
      low_hemoglobin = as.factor(low_hemoglobin), 
      high_hemoglobin = as.factor(high_hemoglobin), 
      hemoglobin_ordered = as.factor(hemoglobin_ordered),
      low_bicarbonate = as.factor(low_bicarbonate), 
      high_bicarbonate = as.factor(high_bicarbonate), 
      bicarbonate_ordered = as.factor(bicarbonate_ordered),
      low_creatinine = as.factor(low_creatinine), 
      high_creatinine = as.factor(high_creatinine), 
      creatinine_ordered = as.factor(creatinine_ordered)
    )

```

##Statistical comparison of HAI to Non-HAI
```{r}
# Remove extreme outliers
Q1 <- quantile(clean_patient$adjusted_icu_los, 0.25, na.rm = TRUE)
Q3 <- quantile(clean_patient$adjusted_icu_los, 0.75, na.rm = TRUE)
IQR_val <- Q3 - Q1

stat_patients <- clean_patient %>%
  filter(adjusted_icu_los >= (Q1 - 1.5*IQR_val) & adjusted_icu_los <= (Q3 + 1.5*IQR_val))
```

```{r}
#Length of stay in hours
stat_patients %>%
  group_by(HAI_flag) %>%
  summarize(
    mean_LOS = mean(adjusted_icu_los, na.rm = TRUE),
    median_LOS = median(adjusted_icu_los, na.rm = TRUE),
    sd_LOS = sd(adjusted_icu_los, na.rm = TRUE),
    n = n()
  )
```

```{r}
#Number of Procedures
stat_patients %>%
  group_by(HAI_flag) %>%
  summarize(
    mean_proc = mean(num_procedures, na.rm = TRUE),
    median_proc = median(num_procedures, na.rm = TRUE),
    sd_proc = sd(num_procedures, na.rm = TRUE),
    n = n()
  )
```

```{r}
#Number of diagnoses
stat_patients %>%
  group_by(HAI_flag) %>%
  summarize(
    mean_diag = mean(num_diagnoses, na.rm = TRUE),
    median_diag = median(num_diagnoses, na.rm = TRUE),
    sd_diag = sd(num_diagnoses, na.rm = TRUE),
    n = n()
  )
```

```{r}
#patient ages
stat_patients %>%
  group_by(HAI_flag) %>%
  summarize(
    mean_age = mean(anchor_age, na.rm = TRUE),
    median_age = median(anchor_age, na.rm = TRUE),
    sd_age = sd(anchor_age, na.rm = TRUE),
    n = n()
  )
```

Summary table for variables
```{r}
# Find all variables that categorical
cat_cols <- clean_patient %>%
  select(where(is.factor)) %>%
  select(-HAI_flag) %>% 
  colnames()

# Find all variables that continuous
cont_cols <- clean_patient %>%
  select(where(is.numeric)) %>% 
  colnames()

# Creat summary table with gtsummary adding appropriate values
summary_tbl <- tbl_summary(
  clean_patient,
  type = list(
    all_continuous() ~ "continuous",
    all_categorical() ~ "categorical"
  ),
  statistic = list(
    all_continuous() ~ "{mean} ({sd})",
    all_categorical() ~ "{n} ({p}%)"
  )
)
```

##Exploratory Data Analysis
Histograms to look for data distribution
```{r, fig.width= 11}
#decreasing the fonts for the titles and axis
common_theme <- theme(plot.title = element_text(size = 11),
                      axis.title.x = element_text(size = 10),
                      axis.title.y = element_text(size = 10)) 

hist_Age <- gf_histogram(~ anchor_age, data = clean_patient, fill = "steelblue", color = "black", bins = 30) %>%
  gf_labs(title = "Age Distribution") + common_theme

hist_los <- gf_histogram(~ adjusted_icu_los, data = clean_patient, fill = "steelblue", color = "black", bins = 30) %>%
  gf_labs(title = "ICU Length of Stay Distribution") + common_theme

hist_diag <- gf_histogram(~ num_diagnoses, data = clean_patient, fill = "steelblue", color = "black", bins = 30) %>%
  gf_labs(title = "Number of Diagnoses Distribution") + common_theme

hist_proc <- gf_histogram(~ num_procedures, data = clean_patient, fill = "steelblue", color = "black", bins = 30) %>%
  gf_labs(title = "Number of Procedures Distribution") + common_theme

```

BoxPlots
```{r}
#decreasing the fonts for the titles
boxplot_theme <- theme(plot.title = element_text(size = 10))
                       
box_age <- gf_boxplot(clean_patient$anchor_age ~ clean_patient$HAI_flag, fill = "cadetblue", color = "black") %>%
  gf_labs(x = NULL, y = NULL, title = "Age") + boxplot_theme
p2 <- gf_boxplot(clean_patient$adjusted_icu_los ~ clean_patient$HAI_flag, fill = "cadetblue", color = "black") %>%
  gf_labs(x = NULL, y = NULL, title = "Length of Stay") + boxplot_theme
p3 <- gf_boxplot(clean_patient$num_diagnoses ~ clean_patient$HAI_flag, fill = "cadetblue", color = "black") %>%
  gf_labs(x = NULL, y = NULL, title = "Number of Diagnoses") + boxplot_theme
p4 <- gf_boxplot(clean_patient$num_procedures ~ clean_patient$HAI_flag, fill = "cadetblue", color = "black") %>%
  gf_labs(x = NULL, y = NULL, title = "Number of Procedures") + boxplot_theme

#Arrange all the boxplots and histograms into a grid
grid.arrange( hist_Age, box_age, hist_los, p2, hist_diag, p3, hist_proc, p4, ncol = 2)
```

Handle right skewness in length of stay and procedures
```{r}
clean_patient$icu_los_log <- log(clean_patient$adjusted_icu_los)

#Some patients have zero procedures need log
clean_patient$num_procedures_log <- log1p(clean_patient$num_procedures)
```

Remove extreme outliers
```{r}
#Find inter Quantile range
Q1 <- quantile(clean_patient$icu_los_log, 0.25)
Q3 <- quantile(clean_patient$icu_los_log, 0.75)
IQR_val <- Q3 - Q1

# Filter within 1.5*IQR
clean_patient <- clean_patient %>%
  filter(icu_los_log >= (Q1 - 1.5*IQR_val) & icu_los_log <= (Q3 + 1.5*IQR_val))
```

Looking at the box plot anchor ages has some outliers in the very young category. This makes sense, as most young people are considered "healthy" and therefore do not spend much time in the ICU. Since there are only about 28 values that are less than the 1st quartile and the distribution seems pretty even, it is decided to not transform these values as they will add useful variation to the model. However there is strong right skewness and many outliers in length of stay and in number of procedures performed. These variable will be transformed using natural log and 1.5*IQR applied to remove extreme outliers.

Proportion of HAI to not HAI
```{r}
table(clean_patient$HAI_flag)

# Proportion of each class
prop.table(table(clean_patient$HAI_flag))
```

Graphs of los and procedures after transformation and outlier removal
```{r}
p2log <- gf_boxplot(clean_patient$icu_los_log ~ clean_patient$HAI_flag, fill = "cadetblue", color = "black") %>%
  gf_labs(x = NULL, y = NULL, title = "Length of Stay log transform") + boxplot_theme
p4log <- gf_boxplot(clean_patient$num_procedures_log ~ clean_patient$HAI_flag, fill = "cadetblue", color = "black") %>%
  gf_labs(x = NULL, y = NULL, title = "Number of Procedures log transform") + boxplot_theme
hist_los_log <- gf_histogram(~ icu_los_log, data = clean_patient, fill = "steelblue", color = "black", bins = 30) %>%
  gf_labs(title = "Log of ICU Length of Stay Distribution") + common_theme
hist_proc_log <- gf_histogram(~ num_procedures_log, data = clean_patient, fill = "steelblue", color = "black", bins = 30) %>%
  gf_labs(title = "Log of Number of Procedures Distribution") + common_theme

grid.arrange(hist_los, hist_los_log, hist_proc, hist_proc_log, hist_Age, hist_diag)

grid.arrange(box_age, p3, p2, p2log, p4, p4log,  ncol = 2)
```

BarPlots for Categorical Variables
```{r}
custom_colors <- scale_fill_manual(values = c("hai" = "darkorange", "not_hai" = "steelblue"))

plot_variable_group <- function(df, vars, group_name) {
  df_long <- df %>%
    select(all_of(vars), HAI_flag) %>%
    pivot_longer(cols = -HAI_flag, names_to = "Variable", values_to = "Value")
   df_long <- df_long %>% mutate(Value = factor(Value))
  
  gf_bar(~ Value | Variable, fill = ~ HAI_flag, position = "dodge", data = df_long) %>%
    gf_labs(title = paste("Distribution of", group_name),
            x = "Value", y = "Count") +
    common_theme + custom_colors +
    facet_wrap(~ Variable, scales = "free_x")
}

demographics <- c("gender")
diagnoses <- c("has_diabetes", "has_ckd", "has_resp", "has_ld", "has_cancer", "has_hd",
               "has_hyperlipidemia", "has_acute_anemia", "has_acidosis",
               "has_nicotine_depend", "has_thrombocytopenia")
procedures <- c("has_cath_bld", "has_surg", "has_feedingtube", "has_bronch_biopsy")
medications <- c("cardiovascular", "electrolytes_fluid", "diuretic", "respiratory",
                 "gi", "other", "antibiotic", "non_therapeutic", "sedative", "steroid",
                 "none","vitamin", "insulin", "analgesic", "vaccine", "neuroleptics",
                 "antiseptic", "glucagon", "neurologic", "volume_expander",
                 "antihistamine", "neuromuscular", "vasopressors", "immunosuppressants",
                 "endocrine", "phosphate_binder")
labs <- c("low_glucose", "high_glucose", "glucose_ordered",
          "low_sodium", "high_sodium", "sodium_ordered",
          "low_potassium", "high_potassium", "potassium_ordered",
          "low_ph_bld", "high_ph_bld", "low_ph_urn", "high_ph_urn", "ph_ordered", 
          "low_chloride", "high_chloride", "chloride_ordered",
          "low_hemoglobin", "high_hemoglobin",  "hemoglobin_ordered",
          "low_bicarbonate", "high_bicarbonate", "bicarbonate_ordered",
          "low_creatinine", "high_creatinine",  "creatinine_ordered")

plot_demo <- plot_variable_group(clean_patient, demographics, "Demographics")
plot_diag <- plot_variable_group(clean_patient, diagnoses, "Diagnoses")
plot_proc <- plot_variable_group(clean_patient, procedures, "Procedures")
plot_med <- plot_variable_group(clean_patient, medications, "Medications")
plot_labs <- plot_variable_group(clean_patient, labs, "Labs")
plot_diag
plot_proc
```

Drop untransformed columns
```{r}
clean_patient <- clean_patient %>%
  select(-adjusted_icu_los,
         -num_procedures)

```

###Inferential Statistics
#Chi-Square Test
```{r, warning=FALSE}
#Create a variable with all categorical variables without predictor
cat_cols <- clean_patient %>%
  select(where(is.factor)) %>%
  select(-HAI_flag) %>% 
  colnames()

# Chi-square tests
# Create a loop to get chi-square for each variable in the categorical data variable
chi_results <- lapply(cat_cols, function(col) {
  tbl <- table(clean_patient[[col]], clean_patient$HAI_flag)
  test <- chisq.test(tbl)
  data.frame(
    variable = col,
    p_value = test$p.value,
    type = "Categorical"
  )
}) %>% bind_rows()
```

#t-test for continuous variables
```{r, warning=FALSE}
#Create a variable for all the continuous variables for t-test
cont_cols <- clean_patient %>%
  select(where(is.numeric)) %>% 
  colnames()

# Loop through all the continuous variables to perfrom t-test
t_results <- lapply(cont_cols, function(col) {
  test <- t.test(clean_patient[[col]] ~ clean_patient$HAI_flag)
  data.frame(
    variable = col,
    p_value = test$p.value,
    type = "Continuous"
  )
}) %>% bind_rows()

#Combine chi-square and t-test results
test_results <- bind_rows(chi_results, t_results)

t_results$variable_clean <- gsub("_log$", "", t_results$variable)

final_tbl <- summary_tbl %>%
  modify_table_body(
    ~ left_join(.x, t_results, by = c("variable" = "variable_clean"))
  )
```

#Analyze results
```{r}
#Varibles with the least statistical significance
test_results %>%
  filter(p_value > 0.05) %>% arrange(p_value)

#Variable with the most statistical significance
test_results %>%
  filter(p_value < 0.05) %>% arrange(p_value)
```

Barplots of top 6 categorical predictors
```{r}
top_cats <- c("has_feedingtube", "has_surg", "antiseptic", "has_cath_bld",
              "neuromuscular", "low_ph_bld")

df_long <- melt(clean_patient, id.vars = "HAI_flag", measure.vars = top_cats)

# Summarize proportions
df_props <- df_long %>%
  group_by(variable, value, HAI_flag) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(prop = count / sum(count))

df_label <- df_props %>% filter(HAI_flag == "hai")

ggplot(df_props, aes(x = factor(value), y = prop, fill = HAI_flag)) +
  geom_col(position = "fill") +
  geom_text(
    data = df_label,
    aes(label = scales::percent(prop, accuracy = 1)),
    position = position_fill(vjust = 0.93),
    color = "white",
    size = 3.75
  ) +
  facet_wrap(~variable, scales = "free_x") +
  scale_fill_manual(values = c("hai" = "darkorange", "not_hai" = "steelblue")) +
  labs(
    title = "Proportion of HAI vs Non-HAI",
    x = "Category Value (0 = No, 1 = Yes)",
    y = "Proportion",
    fill = "HAI Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.05),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

#check for multicollinearity
```{r}
model1 <- glm(HAI_flag ~ ., data = clean_patient, family = binomial)

summary(model1)
```

#Create a cleaned correlation plot
```{r}
#One-hot encode categorical data
dummy_data <- model.matrix(HAI_flag ~ . - 1, data = clean_patient)
corr_dummy <- cor(dummy_data, use = "pairwise.complete.obs")

#create a correlation plot with a cleaner cutoff and group high correlated values together
high_corr_vars <- findCorrelation(corr_dummy, cutoff = 0.3, names = TRUE, verbose = FALSE)
corr_reduced <- corr_dummy[high_corr_vars, high_corr_vars]

# Cluster and plot
corrplot(corr_reduced, 
         method = "color", 
         type = "upper", 
         order = "hclust", 
         tl.cex = 0.6,
         col = colorRampPalette(c("blue", "white", "red"))(200))
```

Group together to remove redundancy and remove unnecessary variables
```{r}
#Remove unnecessary variables and create chem_labs
clean_patient <- clean_patient %>%
  select(-hemoglobin_ordered, -creatinine_ordered, -sodium_ordered, -potassium_ordered,
         -bicarbonate_ordered, -chloride_ordered, -glucose_ordered, -ph_ordered,
         -high_hemoglobin, -high_ph_urn, -low_ph_urn,  -low_glucose, -admission_type)
         
      
```

Create hospital location uniformity and remove rare transfer events
```{r}
#Find out frequency of hospital location and remove rare transfers
clean_patient %>%
  count(admission_location) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  arrange(desc(n))
```

```{r}
#convert back to character to use case when and remove spaces for modeling. 
#Remove low admission locations. Re-factor variable.
clean_patient <- clean_patient %>%
  mutate(admission_location = as.character(admission_location)) %>%
  mutate(admission_location = case_when(
    admission_location == "emergency room" ~ "er",
    admission_location == "transfer from hospital" ~ "transfer",
    admission_location == "physician referral" ~ "dr_referral",
    admission_location == "walk-in/self referral" ~ "self_referral",
    TRUE ~ "other"
  )) %>%
  mutate(admission_location = factor(admission_location,
                                     levels = c("er", "transfer", "dr_referral",
                                                "self_referral", "other")))
```

###FAMD
#Scale data
```{r}
patient_scale <- clean_patient %>%
  mutate(
    anchor_age = scale(anchor_age),
    icu_los_log = scale(icu_los_log),
    num_diagnoses = scale(num_diagnoses),
    num_procedures_log = scale(num_procedures_log)
  ) 
```

```{r}
#Remove HAI_flag and create screeplot
famd_results <- FAMD(patient_scale %>% select(-HAI_flag), graph = FALSE)

fviz_famd_var(famd_results, repel = TRUE)
fviz_screeplot(famd_results)
```

Based on the FAMD variable importance plot remove other unimportant variables.
```{r}
#Remove unnecessary variables and create chem_labs
patient_scale <- patient_scale %>%
  select(-endocrine, -antihistamine, -steroid, -neurologic, -sedative, -vaccine,
         -gi, -electrolytes_fluid, -other, -has_nicotine_depend, -gender)
         
```

Split data
```{r}
RNGkind(sample.kind = "Rejection")
set.seed(1313)
train_index <- createDataPartition(patient_scale$HAI_flag, p = 0.7, list = FALSE)
train_data <- patient_scale[train_index, ]
test_data  <- patient_scale[-train_index, ]

```

No information Rate
```{r}
NIR <- max(table(patient_scale$HAI_flag)) / nrow(patient_scale)
NIR
```

Handle class imbalance using a recipe of downsampling and upsampling
```{r}
#up sampling with mild down sampling
rec_balanced <- recipe(HAI_flag ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = FALSE) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  step_downsample(HAI_flag, under_ratio = 1.5) %>%
  step_upsample(HAI_flag, over_ratio = 1.3)

rec_balanced_prep <- prep(rec_balanced, training = train_data)
balanced_train_data <- juice(rec_balanced_prep)

```


```{r}
#Create the training control using 10-fold Cross-validation and ROC/Sens/Spec evaluation
train_ctrl <- trainControl(
      method = "cv",
      number = 10,
      classProbs = TRUE,
      summaryFunction = twoClassSummary
    )
```

Baseline Full Logistic Regression Model
```{r}
#recipe
fit_LR = train(rec_balanced,
               data = train_data,
               method = "glm",
               trControl = train_ctrl)

fit_LR
```

ROC curve for models:
```{r, fig.width=5, fig.height=5}
# LR Baseline model
probs_LR_pred <- predict(fit_LR, newdata = test_data, type = "prob")[,2]

# ROC Curves
roc_LR <- roc(response = test_data$HAI_flag, predictor = probs_LR_pred)

# Plot ROC Curves
plot.roc(roc_LR, col = "navy", main = "Baseline ROC Logistic Regression Model", cex.main = 0.9, lwd = 2)

# AUC Values
auc_LR <- auc(roc_LR)

# Add AUC Text
text(x = 0.2, y = 0.35, labels = paste("AUC: ", round(auc_LR, 4)), col = "navy", cex = 0.8)
  
```       

Random Forest Parameter Tuning
```{r}
#recipe
RNGkind(sample.kind = "Rejection")
set.seed(1313)

fit_fullRF <- train(
                rec_balanced,
                data = train_data,
                method = "rf",
                tuneGrid = expand.grid(mtry = c(11, 12, 13)),
                ntree = 400,
                trControl = train_ctrl)
fit_fullRF
```

#XGBoost
```{r, warning=FALSE, message=FALSE}
xgbGrid <- expand.grid(nrounds = 400,
                       max_depth = c(2, 3, 4),
                       eta = c(0.01, 0.02, 0.03), 
                       gamma = 0,
                       colsample_bytree = 1,
                       min_child_weight = 1,
                       subsample = 0.8)

fit_xgbo <- train(
               rec_balanced, 
               data = train_data, 
               method = "xgbTree", 
               metric = "roc", 
               trControl = train_ctrl, 
               tuneGrid = xgbGrid
               )

fit_xgbo
```

Artificial Neural Networks
```{r}
RNGkind(sample.kind = "Rejection")
set.seed(1313)
fit_fullannu <- train(rec_balanced,
                 data = train_data, 
                 method = "nnet",
                 tuneGrid = expand.grid(size = c(1, 2), 
                                        decay = seq(0.22, 0.28, by = 0.02)),
                 trace = FALSE,
                 maxit = 400,
                 trControl = train_ctrl)
fit_fullannu
```

Find if ANN model converges:
```{r}
fit_fullannu$finalModel$convergence
```

double cross validation 
```{r}
RNGkind(sample.kind = "Rejection")
set.seed(1313) #Set seed for consistency
n = nrow(patient_scale) #define n on the rows in the dataset
n_outer_folds = 5 #number of folds for double cross validation
# initiate vector as a factor
predicted_Outcome_outer = factor(vector(mode = "character", length = n), 
                                  levels = levels(patient_scale$HAI_flag))

groups = rep(1:n_outer_folds, length = n)
cv_groups = sample(groups, n) # randomization for data in groups

for(ii in 1:n_outer_folds){
  in_test = (cv_groups == ii) # Testing set, logical vector indicating group ii
  in_train = (cv_groups != ii) # Training set, logical vector indicating NOT group ii

  # Training set, logical vector indicating NOT group ii
  
  train_fold <- patient_scale[in_train, ]
  test_data = patient_scale[in_test, ]
  # Add recipe to training fold
  rec_balanced <- recipe(HAI_flag ~ ., data = train_data) %>%
    step_dummy(all_nominal_predictors(), one_hot = FALSE) %>%
    step_center(all_predictors()) %>%
    step_scale(all_predictors()) %>%
    step_downsample(HAI_flag, under_ratio = 1.5) %>%
    step_upsample(HAI_flag, over_ratio = 1.3)
  
  rec_fold <- prep(rec_balanced, training = train_fold)
  dataused <- juice(rec_fold) #Data used in training for cross validation
  test_data_processed <- bake(rec_fold, new_data = test_data)
  
  
  
  ctrl = trainControl(method = "cv", number = 10, classProbs = TRUE,
                      summaryFunction = twoClassSummary) 
  #10-fold cross validation

  #models for cross validation
  fit_lr = train(HAI_flag ~ .,
                  data = dataused,
                  method = "glm",
                  metric = "ROC",
                  trControl = ctrl)
  fit_rf <- train(HAI_flag ~ .,
                  data = dataused, 
                  method = "rf",
                  tuneGrid = expand.grid(mtry = c(11, 12, 13)),
                  ntree = 400,
                  metric = "ROC",
                  trControl = ctrl)
  fit_xgb <- train(HAI_flag ~ ., 
                  data = dataused, 
                  method = "xgbTree", 
                  metric = "ROC", 
                  trControl = ctrl, 
                  tuneGrid = expand.grid(nrounds = 400,
                                         max_depth = c(2, 3, 4),
                                         eta = c(0.01, 0.02, 0.03),
                                         gamma = 0,
                                         colsample_bytree = 1,
                                         min_child_weight = 1,
                                         subsample = 0.8),
                  verbosity = 0)
  fit_ann = train(HAI_flag ~., 
                  data = dataused,
                  method = "nnet",
                  tuneGrid = expand.grid(size = c(1, 2), 
                                         decay = seq(0.22, 0.28, by = 0.02)),
                  trace = FALSE,
                  preProc = c("center", "scale"),
                  maxit = 400,
                  metric = "ROC",
                  trControl = ctrl)
  #pull the maximum ROC after 10 fold cross validation
  all_roc_CV <- c(max(fit_lr$results$ROC),
                  max(fit_rf$results$ROC),
                  max(fit_xgb$results$ROC),
                  max(fit_ann$results$ROC))
  #List of all models in double cross validation                     
  all_train_output <- list(fit_lr, fit_rf, fit_xgb, fit_ann)
  
  #Find the best models with the best accuracy for each fold in outer cross validation
  allbest <- which(all_roc_CV == max(all_roc_CV))
  whichbest <- ifelse(length(allbest)==1, allbest, sample(allbest,1))
  bestmodel_train_output <- all_train_output[[whichbest]]
  bestmodel_name <- c("Logistic Regression Full", 
                      "Random Forest", "XGBoost", "ANN")[whichbest]
  cat("\r Within outer fold", ii, ", the best model", "is", whichbest, "=", bestmodel_name)
  cat('\nAt time', format(Sys.time(),'%H:%M:%S'))  # checking how long to run
  #Use best model for prediction against the test set
  pred_Outcome <- predict(bestmodel_train_output, newdata = test_data_processed)

  predicted_Outcome_outer[in_test] <- pred_Outcome
}

table(predicted_Outcome_outer,patient_scale$HAI_flag) 
#confusion matrix of best model predictions
Accuracy_Outcome_CV = sum(predicted_Outcome_outer == patient_scale$HAI_flag)/n #Accuracy calculation
Accuracy_Outcome_CV

```

Final Model fit:
```{r}
#Get the final best model
bestmodel_train_output$finalModel

#Bake the test data 
test_data_baked <- bake(rec_balanced_prep, new_data = test_data)

#Final model is fitted to the dataset
predictions <- predict(bestmodel_train_output, newdata = test_data_baked)
probs <- predict(bestmodel_train_output, newdata = test_data_baked, type = "prob")

threshold <- 0.4 #Lower prediction threshold 
predictions <- ifelse(probs[,"hai"] >= threshold, "hai", "not_hai")
predictions <- factor(predictions, levels = levels(test_data_baked$HAI_flag))
cm <- confusionMatrix(predictions, test_data_baked$HAI_flag, positive = "hai")
cm
```

Confusion Matrix
```{r}
# Extract the table and convert to data frame 
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Prediction", "Reference", "Freq")
cm_df <- cm_df %>%
  mutate(Correct = ifelse(Prediction == Reference, "Correct", "Incorrect"))

# Heatmap 
ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Correct, alpha = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) + 
  geom_text(aes(label = Freq), color = "black", size = 5) + 
  scale_fill_manual(values = c("Correct" = "steelblue", "Incorrect" = "darkorange")) +
  scale_alpha(range = c(0.4, 1)) + 
  labs( title = "Confusion Matrix Threshold of 0.4", x = "Actual Class", 
        y = "Predicted Class", fill = "Prediction" ) + 
  theme_minimal(base_size = 14) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none")

```

```{r}
#Confusion matrix with sens, spec, ppv, and npv values
sens <- round(cm$byClass["Sensitivity"], 3)
spec <- round(cm$byClass["Specificity"], 3)
ppv  <- round(cm$byClass["Pos Pred Value"], 3)
npv  <- round(cm$byClass["Neg Pred Value"], 3)


ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Correct, alpha = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_manual(values = c("Correct" = "steelblue", "Incorrect" = "darkorange")) +
  scale_alpha(range = c(0.4, 1)) +
  labs(
    title = "Confusion Matrix (Threshold = 0.4)",
    x = "Actual Class",
    y = "Predicted Class"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none") +
  
  annotate("text", x = 1, y = 1,
           label = paste0("Sensitivity: ", sens),
           color = "black", hjust = 0.5, vjust = 1.85, size = 4.0) +
  
  annotate("text", x = 2, y = 2,
           label = paste0("Specificity: ", spec),
           color = "black", hjust = 0.5, vjust = 1.85, size = 4.0) +

  annotate("text", x = 2, y = 1,
           label = paste0("PPV: ", ppv),
           color = "black", hjust = 0.45, vjust = 1.85, size = 4.0) +
  
  annotate("text", x = 1, y = 2,
           label = paste0("NPV: ", npv),
           color = "black", hjust = 0.45, vjust = 1.85, size = 4.0)
```

ROC curve for models:
```{r, fig.width=5, fig.height=5}
# Logistice Regression
probs_lr <- predict(fit_LR, newdata = test_data, type = "prob")[,2]

# Random Forest
probs_rf <- predict(fit_fullRF, newdata = test_data, type = "prob")[,2]

# XGBoost
probs_xgb <- predict(fit_xgbo, newdata = test_data, type = "prob")[,2]

# ANN
probs_ann <- predict(fit_fullannu, newdata = test_data, type = "prob")[,2]

# ROC Curves
roc_lr <- roc(response = test_data$HAI_flag, predictor = probs_lr)
roc_rf <- roc(response = test_data$HAI_flag, predictor = probs_rf)
roc_xgb <- roc(response = test_data$HAI_flag, predictor = probs_xgb)
roc_ann <- roc(response = test_data$HAI_flag, predictor = probs_ann)

# Plot ROC Curves
plot.roc(roc_lr, col = "navy", main = "ROC Curve comparing models", cex.main = 0.9, lwd = 2)
lines(roc_rf, col = "darkgreen", lwd = 2)
lines(roc_ann, col = "darkorange", lwd = 2)
lines(roc_xgb, col = "maroon", lwd = 2)

# AUC Values
auc_lr <- auc(roc_lr)
auc_rf <- auc(roc_rf)
auc_ann <- auc(roc_ann)
auc_xgb <- auc(roc_xgb)

# Add AUC Text
text(x = 0.2, y = 0.25, labels = paste("AUC: ", round(auc_lr, 4)), col = "navy", cex = 0.8)
text(x = 0.2, y = 0.55, labels = paste("AUC: ", round(auc_rf, 4)), col = "darkgreen", cex = 0.8)
text(x = 0.2, y = 0.35, labels = paste("AUC: ", round(auc_ann, 4)), col = "darkorange", cex = 0.8)
text(x = 0.2, y = 0.45, labels = paste("AUC: ", round(auc_xgb, 4)), col = "maroon", cex = 0.8)

# Add Legend
legend("bottomright", legend = c("Random Forest", "XGBoost", "ANN", "Logistic Regression"  ), col = c( "darkgreen", "maroon", "darkorange", "navy" ), lwd = 2, cex = 0.7)
```

Variable importance plot
```{r}
varImp(bestmodel_train_output)
plot(varImp(bestmodel_train_output), top = 20, main = "Variable Importance")

```

Are antiseptic and has_surg correlated?
```{r}
#Look at chi-square test
chisq.test(patient_scale$antiseptic, patient_scale$has_surg)

table(patient_scale$antiseptic, patient_scale$has_surg)
prop.table(table(patient_scale$antiseptic, patient_scale$has_surg))
```

There is statistical association with antiseptics and surgery but not all patient who have antiseptics also had surgery or all surgeries had chlorohexidine. Chlorohexidine can also be used in mouth washes and wound treatment. Will keep both predictors as is in the models.

Partial Dependency Plot
```{R}
patient_df <- as.data.frame(test_data_processed)

proc_plot <- partialPlot(
  x = bestmodel_train_output$finalModel,                   
  pred.data = patient_df,       
  x.var = "num_procedures_log", 
  which.class = "hai",          
  main = "Partial Dependence of Number of Procedures on HAI",
  xlab = "Log of Number of Procedures",
  ylab = "Predicted Probability of HAI"
)
icu_los_plot <- partialPlot(
  x = bestmodel_train_output$finalModel,                   
  pred.data = patient_df,       
  x.var = "icu_los_log", 
  which.class = "hai",          
  main = "Partial Dependence of Length of Stay on HAI",
  xlab = "Log of Number of ICU length of stay",
  ylab = "Predicted Probability of HAI"
)
num_diag_plot <- partialPlot(
  x = bestmodel_train_output$finalModel,                   
  pred.data = patient_df,       
  x.var = "num_diagnoses", 
  which.class = "hai",          
  main = "Partial Dependence of Number of Diagnoses on HAI",
  xlab = "Number of diagnoses",
  ylab = "Predicted Probability of HAI"
)
age_plot <- partialPlot(
  x = bestmodel_train_output$finalModel,                   
  pred.data = patient_df,       
  x.var = "anchor_age", 
  which.class = "hai",          
  main = "Partial Dependence of Age on HAI",
  xlab = "Age",
  ylab = "Predicted Probability of HAI"
)
antiseptics_plot <- partialPlot(
  x = bestmodel_train_output$finalModel,                   
  pred.data = patient_df,       
  x.var = "antiseptic_X1", 
  which.class = "hai",          
  main = "Partial Dependence of the use of Antiseptics on HAI",
  xlab = "Antiseptics",
  ylab = "Predicted Probability of HAI"
)

```

Export Clean patient table so all other code does not need to be execute prior to EDA and model building and evaluation.
```{r}
write_csv(clean_patient, "clean_patient.csv")
```

Export the summary patient table
```{r}
final_tbl %>%
  as_gt() %>%
  gt::gtsave("summarytable.pdf")
```

